WED-SQL (Draft)

### 2016-05-21 doc/paper/main_en.tex
paper: ch5

\par

\par O SGDB PostgreSQL oferece um mecanismo de travas explicitas para aplicações que necessitam exercer um controle mais
refinado sobre os dados acessados. Essas travas podem ser utilizadas tanto a nível de tabelas quanto a nível de tuplas, ou seja,
linhas ou registros das tabelas. No contexto do WED-SQL, apenas as travas a nível de tuplas são utilizadas explicitamente.

\par A trava a nível de tupla mais comumente utilizada é chamada \emph{FOR UPDATE}. Utilizada em conjunto com o comando SQL
\emph{SELECT}, essa trava, de acesso exclusivo, faz com que outras transações concorrentes tenham que esperar ou abortar
ao tentarem acessar a tupla bloqueada. Outro tipo de trava que também pode ser utilizada a nível de tupla são as 
\emph{Advisory Locks}, ou travas semânticas. Para utilizar esse tipo de trava é preciso definir seu significado semântico na aplicação, uma vez que
o PostgreSQL não obriga que sua regra de acesso seja cumprida, ou seja, os registros não são realmente bloqueados.

\par Por exemplo, no WED-server as WED-transitions pendentes são identificadas pelo par (wid,tgid). Esse identificador é
utilizado pelos WED-workers para requisitar uma \emph{Advisory Lock}, que será concedida apenas à primeira transação que
a solicitar para o dado registro. As demais transações que tentarem acessar o mesmo registro não conseguiram obter a trava,
e estarão livres para tentar executar outras WED-transitions pendentes, embora o registro não esteja de fato bloqueado. Vale
lembrar que, obter essa trava para a instância específica que será atualizada, é mandatório para completar a transação com sucesso.

\par Por que são utilizadas \emph{Advisory Locks} ao invés de \emph{FOR UPDATE} no WED-SQL ? Por dois motivos: primeiramente
pelo modo como o PostgreSQL funciona internamente e, em segundo lugar, pela natureza da operação de  transição de estados
no contexto do WED-SQL.

\par O principal motivo para não utilizar travas do tipo \emph{FOR UPDATE} é que as informações sobre
quais tuplas estão bloqueadas são armazenadas em disco, e não em memória RAM. Com isso, é necessário utilizar um plugin
externo ao SGBD para recuperar tais informações, necessárias ao WED-server para realizar o controle do tempo de vida
de cada transação. Por outro lado, as informações sobre \emph{Advisory Locks} podem ser obtidas facilmente, por meio de
uma simples consulta a um catálogo interno do PostgreSQL.

\par Outro motivo para não se utilizar \emph{FOR UPDATE} no controle transacional, é que o registro que seria bloqueado
não é o registro que será atualizado. As travas são obtidas na tabela Job\_Pool, porém as atualizações são executadas na
tabela WED\_flow. Desse modo é mais simples, seguro e eficiente utilizar travas semânticas no WED-server.

>>>>

\par PostgreSQL provides client applications with several explicit locking mechanisms, just in case a more refined control
over the data is required. These locks can be used in either table or row level and are automatically acquired for the DBMS
native operations. In the WED-SQL context though, only row level locks are explicitly employed.

\par The most common exclusive access row level lock is acquired using the SQL expression \emph{FOR UPDATE} by the end of
a \emph{SELECT} command. When a transaction try to lock a row that was previously locked with that lock, it will be either blocked,
and wait until the current transaction ends, or aborted if the expression \emph{NOWAIT} was present after \emph{FOR UPDATE}.
There is however another type of lock that can be used to achieve this same behavior: they are named \emph{Advisory Locks}.

\par Advisory Locks work in a semantic level, that is, they are only useful when they have some semantic meaning in the 
application. Put in other words, the DBMS do not enforce their use and nothing is really locked. It is up to the application
to use then rightly. As an example, in the WED-server, pending WED-transitions siting in the Job\_Pool table are uniquely
identified by the attributes pair (\emph{wid},\emph{tgid}). WED-workers requests Advisory Locks based on this composite key, 
which is only granted only once for each pair until the transaction is over. When a lock is not granted for a WED-work, it
knows that some other "sibling" is currently working on that WED-transition, even tough the record is not really blocked
on the database, and then it is free to look for another task. Once again, the WED-server will blocked transactions not
holding a valid lock.

\par Given that the \emph{FOR UPDATE} approach seems to be the more straight forward way to manage WED-transitions, why
did we decide to stick with Advisory Locks ? Well, one reason is purely technical and is based on how PostgreSQL
works with different types of locks. Another one is due to how the WED-state transitions is carried out by our framework.

\par Regarding the technical aspects of the DBMS, information about \emph{FOR UPDATE} locks are kept on disk instead of
the RAM memory. Because of that, this data does not show up in system catalogs, and can only be obtained in real time
using a third party module (pgrowlocks) that, besides being supplied by PostgreSQL itself, is rather inefficient. On the
other hand, information about Advisory Locks can be easily, and efficiently, obtained directly from system catalogs.

\par On WED-SQL, a WED-transition begins on the Job\_Pool table and ends on the WED\_flow table. So, it would be somewhat
awkward to lock a row on a table, using a FOR UPDATE lock, to update a record on another one. Not only that but it also 
could get messy. Therefore, the semantic locks based solution is simpler to implement, in this context is safer for usage 
and is also a more elegant approach.
